Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2025-07-04 src/docs_paper-jtb/paper.Rmd
Include equation numbers

Section \@ref(sec:naomi) show that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
Because of this, the Naomi model cannot be written as an LGM of the form described in Section \@ref(sec:lgm), but can be written as an ELGM.

@@

Comparison to further Bayesian inference methods could prove beneficial.
Four possibilities stand out as being particularly valuable.
First, there exist other quadrature rules for moderate dimension, such as the Box-Wilson central composite design [@box1992experimental].
It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
Second, rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
This method can be run by using `tmbstan` and setting `laplace = TRUE`.
When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
Third, NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
It may be difficult to implement such algorithms using `TMB`.
Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
Additionally, direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.
Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.

>>>>

Section \@ref(sec:naomi) (e.g. Equations \@ref(eq:weighted-aggregate) to \@ref(eq:weighted-aggregate2)) shows that the Naomi model involves many-to-one mappings of latent field components to the structured additive predictor for each response.
Because of this, the Naomi model cannot be written as an LGM of the form described in Section \@ref(sec:lgm), but can be written as an ELGM.

@@

Comparison to further Bayesian inference methods could prove beneficial.
Four possibilities stand out as being particularly valuable.
First, there exist other quadrature rules for moderate dimension, such as the Box-Wilson central composite design [@box1992experimental].
It would be of interest to compare INLA with a PCA-AGHQ rule to INLA with other such quadrature rules.
Second, rather than use quadrature to integrate the marginal Laplace approximation, an alternative approach is to run HMC [@monnahan2018no; @margossian2020hamiltonian].
This method can be run by using `tmbstan` and setting `laplace = TRUE`.
When run to convergence, inferential error of this method would solely be due to the Laplace approximation, helping to clarify the extent to which the inferential error of INLA is attributable to the quadrature grid.
Third, NUTS is not especially well suited to sampling from Gaussian latent field models like Naomi.
Other MCMC algorithms, such as blocked Gibbs sampling [@geman1984stochastic] or slice sampling [@neal2003slice], could be considered.
It may be difficult to implement such algorithms using `TMB`.
Many MCMC algorithms are implemented and customisable (including, for example, the choice of block structure) within the `NIMBLE` probabilistic programming language (PPL) [@de2017programming].
Direct implementation of Naomi within the Stan PPL could alter the speed by as much as plus or minus 50% [@monnahan2018no] and would give access to other algorithms implemented in Stan such as variational inference [@kucukelbir2015automatic] and Pathfinder [@zhang2022pathfinder].
However, requiring rewriting the Naomi model log-posterior outside of `TMB` would be a substantial downside.
Finally, it would be of substantial interest to implement the Naomi model using the iterative INLA method via `inlabru`.
However, as `inlabru`, like `R-INLA`, is based on a formula interface, it may not be possible to do so directly.
